---
title: "process_SJMB12_samples"
output: html_document
date: "2025-10-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# setup
```{r}
.libPaths(file.path(Sys.getenv("HOME"),"R/rstudio/4.3.1"))
library(ArchR, lib.loc = file.path(Sys.getenv("HOME"),"R/rstudio/4.3.1")) # seurat 4
library(dplyr)
# known seed for replication
set.seed(1)


workDir <- file.path(Sys.getenv("COMMONR"),"sex_bias")
inpath <- file.path(workDir, "scripts/UBCenrichedE16Multiome/snATAC_mutiome") # dir with original arrow files
outpath <- file.path(workDir, "processed/mouse/multiome/snATAC/UBCenrichedE16Multiome")

set.seed(1)

cat("# Add ArchR params\n")
addArchRThreads(threads = 1) 


```

# build mm39 genome
```{r}


library(BSgenome.Mmusculus.UCSC.mm39)
library(org.Mm.eg.db)
library(GenomicFeatures)
library(GenomicRanges)

gtf_path <- file.path(Sys.getenv("HOME"),"refdata-cellranger-arc-GRCm39-2024-A/genes/genes.gtf.gz")

TxDb.Mmusculus <- makeTxDbFromGFF(
    file = gtf_path,
    format = "gtf",
    organism = "Mus musculus"
)

geneAnno <- createGeneAnnotation(
    TxDb = TxDb.Mmusculus,  
    OrgDb = org.Mm.eg.db
)

# blacklist downloaded from: https://dozmorovlab.github.io/excluderanges/#excluderanges-genomic-ranges-of-problematic-genomic-regions
bed <- read.table(file.path(Sys.getenv("HOME"),"references/mm39.excluderanges.bed"), 
                  sep = "\t", 
                  header = FALSE,
                  colClasses = c("character", "integer", "integer", "integer", "character", "character"))

blacklist <- GRanges(
    seqnames = bed[,1],
    ranges = IRanges(start = bed[,2] + 1, end = bed[,3])
)

genomeAnno <- createGenomeAnnotation(
    genome = BSgenome.Mmusculus.UCSC.mm39,
    blacklist = blacklist
)

saveRDS(geneAnno, file.path(Sys.getenv("HOME"),"references/Mus_musculus/mm39_ArchR/mm39_geneAnnotation.rds"))
saveRDS(genomeAnno, file.path(Sys.getenv("HOME"),"references/Mus_musculus/mm39_ArchR/mm39_genomeAnnotation.rds"))

```


# Create ArchR project
```{r}


script_content <- '
library(ArchR)
library(dplyr)
library(rhdf5)

cat("# Set seed for replication\n")
set.seed(1)

cat("# Add ArchR params\n")
addArchRThreads(threads = 6) 

geneAnno <- readRDS(file.path(Sys.getenv("HOME"),"references/Mus_musculus/mm39_ArchR/mm39_geneAnnotation.rds"))
genomeAnno <- readRDS(file.path(Sys.getenv("HOME"),"references/Mus_musculus/mm39_ArchR/mm39_genomeAnnotation.rds"))

# Define paths
workDir <- file.path(Sys.getenv("COMMONR"),"sex_bias")
common_dir <- file.path(workDir, "processed/mouse/multiome", "cellranger-arc") # mapped files (fragment files)
inpath <- file.path(workDir, "scripts/UBCenrichedE16Multiome/snATAC_mutiome") # dir with original arrow files
outpath <- file.path(workDir, "processed/mouse/multiome/snATAC/UBCenrichedE16Multiome")

# Create output directories if they dont exist
dir.create(outpath, showWarnings = FALSE, recursive = TRUE)
dir.create(inpath, showWarnings = FALSE, recursive = TRUE)

# Sample information
unique_paths <- c(
  "female-e16",
  "male-e16"
)
cat(paste0("# Total samples: ", length(unique_paths), "\n"))

sample_names <- c(  "female-e16",
  "male-e16")
cat(paste0("# Total sample names: ", length(sample_names), "\n"))

# Full paths to sample directories
sample_dirs <- file.path(common_dir, unique_paths)
names(sample_dirs) <- sample_names


# Function to get barcodes from filtered h5 matrix
getValidBarcodesFromFilteredH5 <- function(dir, sample_name) {
  h5_file <- file.path(dir, "outs/filtered_feature_bc_matrix.h5")
  
  if (!file.exists(h5_file)) {
    cat(paste0("Warning: H5 file not found at ", h5_file, "\n"))
    return(NULL)
  }
  
  barcodes <- h5read(h5_file, "matrix/barcodes")
  
  # Store barcodes in a named list
  result <- list(barcodes = barcodes)
  names(result) <- sample_name
  return(result)
}

# Loop through each sample and create Arrow files with error handling
for (i in seq_along(sample_dirs)) {
  
  sample <- names(sample_dirs)[i]
  sample_dir <- sample_dirs[i]
  
  cat(paste0("\n# Processing sample: ", sample, " (", i, "/", length(sample_dirs), ")\n"))
  
  # Get fragment file
  fragment_file <- file.path(sample_dir, "outs/atac_fragments.tsv.gz")
  
  if (!file.exists(fragment_file)) {
    cat(paste0("Warning: Fragment file not found for ", sample, ". Skipping.\n"))
    next
  }
  
  inputFiles <- fragment_file
  names(inputFiles) <- sample
  
  cat(paste0("# Fragment file: ", fragment_file, "\n"))
  
  cat("# Get cells with Cell Ranger validated barcodes\n")
  CellRangerValidatedBarcodes <- getValidBarcodesFromFilteredH5(sample_dir, sample)
  
  if (is.null(CellRangerValidatedBarcodes)) {
    cat(paste0("Warning: Could not get validated barcodes for ", sample, ". Skipping.\n"))
    next
  }
  
  validateNum <- length(CellRangerValidatedBarcodes[[sample]])
  cat(paste0("# Keeping only ", validateNum, " cells validated by Cell Ranger for ", sample, "\n"))
  
  # Try-Catch block for createArrowFiles
  tryCatch({
    ArrowFile <- createArrowFiles(
      inputFiles = inputFiles,
      sampleNames = sample,
      minTSS = 0,
      minFrags = 0,  # Filter later
      addTileMat = TRUE,
      addGeneScoreMat = TRUE,
            geneAnnotation = geneAnno,
    genomeAnnotation = genomeAnno,
      validBarcodes = CellRangerValidatedBarcodes,
      force = TRUE
    )
    cat(paste0("# Successfully created Arrow file for ", sample, "\n"))
  }, error = function(e) {
    cat(paste0("Error creating Arrow file for ", sample, ": ", e$message, "\n"))
  })
  
} # End of the loop

cat("\n# Processing completed for all samples.\n")

# Get all Arrow files
ArrowFiles <- list.files(inpath, pattern = ".arrow$", full.names = TRUE)
cat(paste0("# Number of arrow files: ", length(ArrowFiles), "\n"))

# Create ArchR Project
cat("# Creating ArchR Project\n")
projPat1 <- ArchRProject(
  ArrowFiles = ArrowFiles,
  outputDirectory = inpath,
      geneAnnotation = geneAnno,
    genomeAnnotation = genomeAnno,
  copyArrows = TRUE # Recommended so that you have an original copy
)

cat(paste0("Memory Size = ", round(object.size(projPat1) / 10^6, 3), " MB\n"))

cat("# Saving ArchR project\n")
saveArchRProject(ArchRProj = projPat1, outputDirectory = outpath, load = FALSE)

cat("# Project saved successfully!\n")
'


# Write to the correct location with correct filename
writeLines(script_content, con = file.path(inpath, "save_arrowFiles_to_RDS.R"))
file.path(inpath, "save_arrowFiles_to_RDS.R")
```

```{bash}

cd $COMMONR/sex_bias/scripts/UBCenrichedE16Multiome/snATAC_mutiome

bsub -q priority -P save_arrowFiles_to_RDS -R "rusage[mem=30GB]" -R "span[hosts=1]" -n 6 -o "log/inferDoublets.%J.out" -e "log/inferDoublets.%J.out" "source activate R_conda_env_4.0.5; Rscript save_arrowFiles_to_RDS.R"
```

# process snRNA from matched samples
```{r}


library(anndata)

rna_h5ad <- anndata::read_h5ad(file.path(Sys.getenv("COMMONR"),'sex_bias/processed/mouse/multiome/writes/UBCenrichedE16Multiome_scVI.h5ad'))

counts_t <- Matrix::t(rna_h5ad[['layers']][['counts']])
rna_obj <- Seurat::CreateSeuratObject(
  counts = counts_t,
  meta.data = rna_h5ad[['obs']],
  project = "SJMB12_RNA",
  min.cells = 0,
  min.features = 0
)

# Set the data slot
log1p_norm <- Matrix::t(rna_h5ad[['layers']][['log1p_norm']])
rna_obj <- Seurat::SetAssayData(
  object = rna_obj,
  slot = "data",
  new.data = log1p_norm
)
rna_obj@meta.data
umap_embedding <- rna_h5ad[['obsm']][['X_umap']]
colnames(umap_embedding) <- paste0("UMAP_", 1:ncol(umap_embedding))
rownames(umap_embedding) <- rownames(rna_h5ad[['obs']])

rna_obj[["umap"]] <- Seurat::CreateDimReducObject(
  embeddings = umap_embedding,
  key = "UMAP_",
  assay = "RNA"
)

pca_embedding <- rna_h5ad[['obsm']][['X_pca']]
colnames(pca_embedding) <- paste0("PC_", 1:ncol(pca_embedding))
rownames(pca_embedding) <- colnames(rna_obj)
rna_obj[["pca"]] <- Seurat::CreateDimReducObject(
  embeddings = pca_embedding,
  key = "PC_",
  assay = "RNA"
)

scVI_embedding <- rna_h5ad[['obsm']][['X_scVI']]
colnames(scVI_embedding) <- paste0("scVI_", 1:ncol(scVI_embedding))
rownames(scVI_embedding) <- colnames(rna_obj)
rna_obj[["scVI"]] <- Seurat::CreateDimReducObject(
  embeddings = scVI_embedding,
  key = "scVI_",
  assay = "RNA"
)

scANVI_embedding <- rna_h5ad[['obsm']][['X_scANVI']]
colnames(scANVI_embedding) <- paste0("scANVI_", 1:ncol(scANVI_embedding))
rownames(scANVI_embedding) <- colnames(rna_obj)
rna_obj[["scANVI"]] <- Seurat::CreateDimReducObject(
  embeddings = scANVI_embedding,
  key = "scANVI_",
  assay = "RNA"
)

Seurat::DimPlot(rna_obj, group.by  = 'human_celltype_pred_broad')
Seurat::FeaturePlot(rna_obj, features = c('Lmx1a'))

saveRDS(rna_obj,file.path(Sys.getenv("COMMONR"),"sex_bias/processed/mouse/multiome/writes/UBCenrichedE16Multiome_seurat.RDS"))


```

# START HERE: Start from project RDS

# Filter snATAC
```{r}
workDir <- file.path(Sys.getenv("COMMONR"),"sex_bias")
outpath <- file.path(workDir, "processed/mouse/multiome/snATAC/UBCenrichedE16Multiome")

projPat1 <- loadArchRProject(path =outpath) 
getOutputDirectory(projPat1)

nfrag=1000
cat(paste0("# filtering based on nFragments: ",nfrag, "."))
idxPass <- which(projPat1$nFrags >= nfrag)
length(idxPass)
cellsPass <- projPat1$cellNames[idxPass]
projPat1 <- projPat1[cellsPass, ]

tssEnrich=4 # this was based on unfiltered distribution
cat(paste0("# filtering based on TSS enrichment: ",tssEnrich, ".\n"))
idxPass <- which(projPat1$TSSEnrichment >= tssEnrich)
length(idxPass)
cellsPass <- projPat1$cellNames[idxPass]
projPat1 <- projPat1[cellsPass, ]

# Estimate doublet scores
cat("# Estimating doublet scores\n")
doubScores <- addDoubletScores(
  projPat1
)

cat(paste0("# filtering doublets with default filter ratio of 1\n"))
projPat1 <- filterDoublets(projPat1)# default filter ratio is 1
projPat1

saveArchRProject(ArchRProj = projPat1, outputDirectory =outpath, load = FALSE)

```

### -- match RNA and ATAC cells

###### --- reset to raw
```{r}


# ArrowFiles <- list.files(inpath, pattern = ".arrow$", full.names = TRUE)
# cat(paste0("# Number of arrow files: ", length(ArrowFiles), "\n"))
# 
# # Create ArchR Project
# cat("# Creating ArchR Project\n")
# projPat1 <- ArchRProject(
#   ArrowFiles = ArrowFiles,
#   outputDirectory = outpath,
#   copyArrows = TRUE # Recommended so that you have an original copy
# )
# saveArchRProject(ArchRProj = projPat1, outputDirectory = outpath, load = FALSE)


```

### -- get RNA and ATAC object
```{r}


# load object from inpath if need to go back to the original obj
projPat1 <- loadArchRProject(path =outpath) 
getOutputDirectory(projPat1)

cat(paste0("Memory Size = ", round(object.size(projPat1) / 10^6, 3), " MB."))

rna_obj <- readRDS(file.path(Sys.getenv("COMMONR"),"sex_bias/processed/mouse/multiome/writes/UBCenrichedE16Multiome_seurat.RDS"))
# rna_obj@assays$RNA$data

```
### --- get shared cells
```{r}

# rename rna cells in ATAC format
sample_meta <- rna_obj@meta.data
sample_meta$atac_cellid <- mapply(function(x,y){ 
  s <- unlist(strsplit(x, split = '_',fixed=T))
  batch_names <- s[2]
  cellname <- s[1]
  sample_id <- y
  paste0(batch_names, '#', cellname , sample_id)
  },
  rownames(sample_meta),sample_meta$sample_batch)
rna_obj@meta.data$atac_cellid <- sample_meta$atac_cellid



# shared cells between RNA and ATAC
shared_cells <- intersect(sample_meta$atac_cellid, rownames(projPat1@cellColData))

length(sample_meta$atac_cellid) # n rna cells
length(rownames(projPat1@cellColData)) # n atac cells
length(shared_cells)

```

### --- subset objects
```{r}

atac_meta_subset <- projPat1@cellColData[shared_cells,]

# merge metadata
cellColData_df <- left_join(as.data.frame(atac_meta_subset) %>% tibble::rownames_to_column("atac_cellid"), sample_meta, by ="atac_cellid")
rownames(cellColData_df) <- cellColData_df$atac_cellid

# subset RNA object
rna_obj_shared <- rna_obj[,rownames(rna_obj@meta.data[rna_obj@meta.data$atac_cellid %in% shared_cells,])]

# subset ATAC object
projPat1_subset <- projPat1[shared_cells, ]

for(col in c('human_celltype_pred', 'human_celltype_pred_broad', 'human_celltype_pred_broad1', 'human_celltype_pred_lvl1', 'human_celltype_pred_lvl2', 'human_celltype_pred_lvl3', 'Age_days', 'Sex', 'leiden', 'Age')){
  projPat1_subset@cellColData[paste0('rna_',col)] <- cellColData_df[rownames(projPat1_subset@cellColData),][col] 

}


cat("# Save project to a RDS.")
saveRDS(rna_obj_shared,file.path(Sys.getenv("COMMONR"),"sex_bias/processed/mouse/multiome/writes/UBCenrichedE16Multiome_ATACshared_seurat.RDS"))

outpath <- file.path(Sys.getenv("COMMONR"),"sex_bias/processed/mouse/multiome/snATAC/UBCenrichedE16Multiome_RNAATACshared")
projPat1_subset@projectMetadata$outputDirectory <- outpath
saveArchRProject(ArchRProj = projPat1_subset, outputDirectory =outpath, load = T)
projPat1_subset
# projPat1_subset@cellColData
```

# add matching RNA cells to ATAC project
```{r}

rna_obj_shared <- readRDS(file.path(Sys.getenv("COMMONR"),"sex_bias/processed/mouse/multiome/writes/UBCenrichedE16Multiome_ATACshared_seurat.RDS"))

library(GenomicFeatures)
library(GenomicRanges)

# Use the SAME GTF that you used for ArchR
gtf_path <- file.path(Sys.getenv("HOME"),"refdata-cellranger-arc-GRCm39-2024-A/genes/genes.gtf.gz")

# Import genes from this GTF
txdb <- makeTxDbFromGFF(gtf_path, format = "gtf")
genes <- genes(txdb)

library(ensembldb)
# Import GTF with gene_name field
gtf <- rtracklayer::import(gtf_path)
gene_info <- unique(as.data.frame(mcols(gtf)[, c("gene_id", "gene_name")]))
gene_info <- gene_info[!is.na(gene_info$gene_name),]

# Match and add gene_name to genes
genes$gene_name <- gene_info$gene_name[match(genes$gene_id, gene_info$gene_id)]

gene_names <- rownames(rna_obj_shared@assays$RNA$counts)
gene_match_idx <- match(gene_names, genes$gene_name)
genes_keep <- !is.na(gene_match_idx)

# Subset to matched genes only
gene_names_matched <- gene_names[genes_keep]
counts_matched <- rna_obj_shared@assays$RNA$counts[genes_keep, ]
genes_matched <- genes[gene_match_idx[genes_keep]]
names(genes_matched) <- gene_names_matched

# Create RangedSummarizedExperiment
rna_meta <- rna_obj_shared@meta.data
rownames(rna_meta) <- rna_meta$atac_cellid
counts_reordered <- rna_obj_shared@assays$RNA$counts
colnames(counts_reordered) <- rna_meta$atac_cellid

# Make sure everything has the same dimensions
counts_matched <- counts_matched[names(genes_matched), ]  # Reorder rows to match genes_matched
colnames(counts_matched) <- rna_meta$atac_cellid

# Verify dimensions match
stopifnot(nrow(counts_matched) == length(genes_matched))
stopifnot(ncol(counts_matched) == nrow(rna_meta))

# Then create the SummarizedExperiment
seRNA <- SummarizedExperiment(
    assays = list(counts = counts_matched),
    rowRanges = genes_matched,
    colData = rna_meta
)

projPat1_subset <- addGeneExpressionMatrix(input = projPat1_subset, seRNA = seRNA, strictMatch = TRUE, force = TRUE)

saveArchRProject(ArchRProj = projPat1_subset, outputDirectory =outpath, load = T)
getAvailableMatrices(projPat1_subset)
```

```{r}

workDir <- file.path(Sys.getenv("COMMONR"),"sex_bias")
outpath <- file.path(workDir, "processed/mouse/multiome/snATAC/UBCenrichedE16Multiome_RNAATACshared")

projPat1 <- loadArchRProject(path =outpath) 

```

# Dim Reduction

```{r}


script_content <- '
library(ArchR)

cat("# known seed for replication\n")
set.seed(1)

cat("# add ArchR params")
addArchRThreads(threads = 1) 

geneAnno <- readRDS(file.path(Sys.getenv("HOME"),"references/Mus_musculus/mm39_ArchR/mm39_geneAnnotation.rds"))
genomeAnno <- readRDS(file.path(Sys.getenv("HOME"),"references/Mus_musculus/mm39_ArchR/mm39_genomeAnnotation.rds"))

workDir <- file.path(Sys.getenv("COMMONR"),"sex_bias")
outpath <- file.path(workDir, "processed/mouse/multiome/snATAC/UBCenrichedE16Multiome_RNAATACshared")

projPat1 <- loadArchRProject(path =outpath) 
getOutputDirectory(projPat1)

cat("# add LSI dimension reduction\n")
projPat1 <- addIterativeLSI(
    ArchRProj = projPat1,
    useMatrix = "TileMatrix",
    name = "IterativeLSI",
    iterations = 2,
    clusterParams = list(resolution = c(2), sampleCells = 10000, maxClusters = 10, n.start= 10),
    varFeatures = 25000,
    dimsToUse = 1:30,
    excludeChr = c(),
     UMAPParams = list(n_neighbors = 40, min_dist = 0.2, metric = "cosine", verbose =
    FALSE, fast_sgd = TRUE, spread = 2)
)


cat("# compute clusters based on LSI\n")

library(reticulate)
use_python(file.path(Sys.getenv("HOME"),"/miniconda3/envs/seurat4.4.0/bin/python"), required = TRUE)
py_config()

projPat1 <- addClusters(
    input = projPat1,
    reducedDims = "IterativeLSI",
    method = "Seurat",
    name = "Clusters",
    algorithm = 4, # leiden
    force = T,
    resolution =  2,
    maxClusters = 100
)

fig_path <- file.path(outpath, "figures")
if (!file.exists(fig_path)) {
  dir.create(fig_path, recursive = TRUE)
  message("Directory created: ", fig_path)
} 

# which sample is in which cluster
cM <- confusionMatrix(paste0(projPat1$Clusters), paste0(projPat1$Sample))

cM <- cM / Matrix::rowSums(cM)

# pdf(file.path(outpath, "figures", "confusion_matrix_sample_vs_cluster.pdf"))
# pheatmap::pheatmap(
#     mat = as.matrix(cM),
#     color = paletteContinuous("whiteBlue"),
#     border_color = "black"
# )
# dev.off()

cat("# add UMAP\n")
projPat1 <- addUMAP(
    ArchRProj = projPat1,
    reducedDims = "IterativeLSI",
    name = "UMAP_ATAC",
    nNeighbors = 40,
    minDist = 0.4,
    metric = "cosine"
)

cat("# add imputated expression.\n")
projPat1 <- addImputeWeights(projPat1)


cat("# saveRDS")
saveArchRProject(ArchRProj =projPat1, outputDirectory =outpath, load = FALSE)
'


writeLines(script_content, con = file.path(inpath, "get_LSI_DimRed_UMAP.R"))
file.path(inpath, "get_LSI_DimRed_UMAP.R")

```

```{bash}

cd $COMMONR/sex_bias/scripts/UBCenrichedE16Multiome/snATAC_mutiome

bsub -q priority -P get_LSI_DimRed_UMAP -R "rusage[mem=100GB]" -R "span[hosts=1]" -n 1 -o "log/get_LSI_DimRed_UMAP.%J.out" -e "log/get_LSI_DimRed_UMAP.%J.out" "source activate R_conda_env_4.0.5; Rscript get_LSI_DimRed_UMAP.R"


```

# Explore UMAP

```{r}

workDir <- file.path(Sys.getenv("COMMONR"),"sex_bias")
inpath <- file.path(workDir, "scripts/UBCenrichedE16Multiome/snATAC_mutiome") # dir with original arrow files
outpath <- file.path(workDir, "processed/mouse/multiome/snATAC/UBCenrichedE16Multiome_RNAATACshared")

proj1 <- loadArchRProject(path = outpath)
proj1

```

```{r}

proj1@cellColData

p1 <- plotEmbedding(ArchRProj = proj1, colorBy = "cellColData", name = "Sample", embedding ="UMAP_ATAC")
p1

p2 <- plotEmbedding(ArchRProj = proj1, colorBy = "cellColData", name = "rna_human_celltype_pred", embedding = "UMAP_ATAC")
p2

p3 <- plotEmbedding(ArchRProj = proj1, colorBy = "cellColData", name = "rna_human_celltype_pred_broad", embedding = "UMAP_ATAC")
p3

metadata_col_4 <- "rna_leiden"
p4 <- plotEmbedding(ArchRProj = proj1, colorBy = "cellColData", name = metadata_col_4, embedding = "UMAP_ATAC")
p4

metadata_col_5 <- "Clusters"
p5 <- plotEmbedding(ArchRProj = proj1, colorBy = "cellColData", name = metadata_col_5, embedding = "UMAP_ATAC")
p5


plotPDF(p1, p2,p3,p4, p5,name = "Plot-UMAP-Metadata.pdf", ArchRProj = proj1, addDOC = FALSE, width = 15, height = 15)
```


# Call peaks with pseudobulks
```{r}
script_content <- '
library(ArchR)

cat("# known seed for replication\n")
set.seed(1)

cat("# add ArchR params")
addArchRThreads(threads = 8) 

geneAnno <- readRDS(file.path(Sys.getenv("HOME"),"references/Mus_musculus/mm39_ArchR/mm39_geneAnnotation.rds"))
genomeAnno <- readRDS(file.path(Sys.getenv("HOME"),"references/Mus_musculus/mm39_ArchR/mm39_genomeAnnotation.rds"))

library(BSgenome.Mmusculus.UCSC.mm39)

workDir <- file.path(Sys.getenv("COMMONR"),"sex_bias")
outpath <- file.path(workDir, "processed/mouse/multiome/snATAC/UBCenrichedE16Multiome_RNAATACshared")

cat("# read in RAW RDS\n")
proj1 <- loadArchRProject(path = outpath) # read from copied arrowfiles
proj1

cat("# pseudo-bulk replicates based on \n")
groupby= "Clusters"
proj1 <- addGroupCoverages(ArchRProj = proj1, groupBy = groupby,  threads = 8)


pathToMacs2 <- file.path(Sys.getenv("HOME"),"miniconda3/envs/R_conda_env_4.0.5/bin/macs2")
genome_size <- sum(seqlengths(proj1@genomeAnnotation$chromSizes))
proj1 <- addReproduciblePeakSet(

    ArchRProj = proj1,

    groupBy = groupby,

    pathToMacs2 = pathToMacs2,
    geneAnnotation = geneAnno,
    genomeAnnotation = genomeAnno,
    genomeSize = "mm"
    
)

proj1 <- addPeakMatrix(proj1)

cat("# saveRDS")
saveArchRProject(ArchRProj = proj1, outputDirectory =outpath, load = FALSE)
'

writeLines(script_content, con= file.path(inpath,   "pseudoBulk_replicates.R"))
file.path(inpath,   "pseudoBulk_replicates.R")
```

```{bash}
cd $COMMONR/sex_bias/scripts/UBCenrichedE16Multiome/snATAC_mutiome

bsub -q priority -P pseudoBulk_replicates_callPeaks -R "rusage[mem=100GB]" -R "span[hosts=1]" -n 1 -o "log/pseudoBulk_replicates_callPeaks.%J.out" -e "log/pseudoBulk_replicates_callPeaks.%J.out" "source activate R_conda_env_4.0.5; Rscript  pseudoBulk_replicates.R"


```

# Explore Peak callled project
```{r}


workDir <- file.path(Sys.getenv("COMMONR"),"sex_bias")
inpath <- file.path(workDir, "scripts/UBCenrichedE16Multiome/snATAC_mutiome") # dir with original arrow files
outpath <- file.path(workDir, "processed/mouse/multiome/snATAC/UBCenrichedE16Multiome_RNAATACshared")


set.seed(1)

cat("# Add ArchR params\n")
addArchRThreads(threads = 1) 

```

```{r}

proj1 <- loadArchRProject(path = outpath)
proj1

colnames(proj1@cellColData)

proj1@cellColData$Sex_celltypeBroad <- paste0(proj1@cellColData$rna_Sex, '_', proj1@cellColData$rna_human_celltype_pred_broad)

proj1@cellColData$Sex_celltype <- paste0(proj1@cellColData$rna_Sex, '_', proj1@cellColData$rna_human_celltype_pred)
```


```{r}
getAvailableMatrices(proj1)

```

# Markers per sex at all celltypes
```{r}

selected_group <- names(table(proj1@cellColData$Sex_celltypeBroad )[table(proj1@cellColData$Sex_celltypeBroad ) > 10])

markerPeaks <- getMarkerFeatures(
    ArchRProj = proj1,
    useMatrix = "PeakMatrix",
    groupBy = "Sex_celltypeBroad",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon",
  useGroups = selected_group,
  bgdGroups = selected_group,
)

saveRDS(markerPeaks, file.path(outpath, 'RDS', 'Sex_celltypeBroad_markerPeaks.RDS'))


```

```{r}

a <- unique(proj1@cellColData["Sex_celltypeBroad"]$Sex_celltypeBroad)


p <- plotBrowserTrack(
    ArchRProj = proj1, 
    groupBy = "Sex_celltypeBroad", 
    geneSymbol = 'Maz',
    upstream = 10000,
    downstream = 10000,
    useGroups = a[grepl('UBC|RL',a, fixed = F)]
    
)



grid::grid.newpage()
grid::grid.draw(p[[1]])


```

### --- overall sex-baised peaks across celltypes
```{r}

markerPeaks <- readRDS(file.path(outpath, 'RDS', 'Sex_celltypeBroad_markerPeaks.RDS'))
# 
# markerList <- getMarkers(markerPeaks, cutOff = "FDR <= 0.05 & Log2FC >= 1")
# names(markerList)

# Get the groups you want
groups_to_plot <- c("M_UBC_GNP_progenitors", "F_UBC_GNP_progenitors", "M_UBC", "F_UBC")

# Subset the SummarizedExperiment
seMarker_subset <- markerPeaks[, colnames(markerPeaks) %in% groups_to_plot]



seMarker_subset@assays@data$AUC
seMarker_subset@assays@data$MeanBGD

heatmapPeaks <- plotMarkerHeatmap(
  seMarker = seMarker_subset, 
  cutOff = "FDR <= 0.01 & Log2FC >= 3 & MeanBGD > 0",
  transpose = F,
  labelRows = F

)
heatmapPeaks 
plotPDF(heatmapPeaks , name = "UBC_Sex_markerPeaks", width = 5, height = 5, ArchRProj = proj1, addDOC = FALSE)


# Loop through each celltype and export as BED file
for (ct in names(markerList)) {
  # Filter data for this celltype
  ct_data <- as.data.frame(markerList[[ct]]) %>%
    subset(FDR < 0.05 & Log2FC > 1) %>%
    mutate(pi = Log2FC*-log10(FDR)) %>% 
     # slice_max(pi, n= 500) %>%
    dplyr::select(seqnames, start, end) %>%
    arrange(seqnames, start) 
    
  
  # Create filename (sanitize celltype name for filename)
  filename <- file.path(outpath, "tables",
                        paste0( ct, ".bed"))
  
  # Write BED file (tab-separated, no header, no quotes, no row names)
  write.table(ct_data, 
              file = filename,
              sep = "\t",
              quote = FALSE,
              row.names = FALSE,
              col.names = FALSE)
  
  cat("Exported", nrow(ct_data), "regions for", ct, "to", filename, "\n")
}



```
### --- Motif enrichment
```{r}

if("Motif" %ni% names(proj1@peakAnnotation)){
    proj1 <- addMotifAnnotations(ArchRProj = proj1, motifSet = "cisbp", name = "Motif")
}

motifsUp <- peakAnnoEnrichment(
    seMarker = seMarker_subset ,
    ArchRProj = proj1,
    peakAnnotation = "Motif",
    cutOff = "FDR <= 0.05 & Log2FC >= 1 & MeanBGD > 0",
    background = "bgdPeaks"
  )


heatmapEM <- plotEnrichHeatmap(motifsUp, n = 10, transpose = TRUE, cutOff = 20, clusterCols = T)

ComplexHeatmap::draw(heatmapEM, heatmap_legend_side = "bot", annotation_legend_side = "bot")
```
### -- pairwise sex-biased 
```{r}


markerPeaks <- getMarkerFeatures(
    ArchRProj = proj1,
    useMatrix = "PeakMatrix",
    groupBy = "Sex_celltype",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon",
  #   useGroups = 'M_UBC_GNP_progenitors',
  # bgdGroups = 'F_UBC_GNP_progenitors',
      useGroups = 'M_UBC_early',
  bgdGroups = 'F_UBC_early',
)

# saveRDS(markerPeaks, file.path(outpath, 'RDS', 'Sex_celltypeBroad_UBC_GNP_progenitors_markerPeaks.RDS'))

```


```{r}
# markerPeaks <- readRDS(file.path(outpath, 'RDS', 'Sex_celltypeBroad_UBC_GNP_progenitors_markerPeaks.RDS'))


# markerList <- getMarkers(markerPeaks, cutOff = "FDR <= 0.05 & abs(Log2FC) >= 1")
# names(markerList)
# 
# marker_df <- markerList$M_UBC_GNP_progenitors %>% subset(FDR < 0.05 & Log2FC > 1)

marker_df <- data.frame(
    markerPeaks@elementMetadata
)


marker_df$FDR <-(markerPeaks@assays@data$FDR$x)
marker_df$Log2FC<- markerPeaks@assays@data$Log2FC$x
marker_df$MeanBGD <- markerPeaks@assays@data$MeanBGD$x
marker_df$MeanDiff <- markerPeaks@assays@data$MeanDiff$x
marker_df$AUC <- markerPeaks@assays@data$AUC$x
marker_df

# marker_df <- marker_df[marker_df$MeanBGD > 0.1,]

# quantile(marker_df$AUC)

marker_df$label <- ifelse(marker_df$FDR < 0.05 & marker_df$Log2FC > 1, 'up_in_M', 
                          ifelse(marker_df$FDR < 0.05 & marker_df$Log2FC < -1 , 'up_in_F', 'NS'))

p1 <- ggplot(marker_df, aes(x = Log2FC, y = -log10(FDR)))+  # log10, not log
  geom_point(aes(color = label))+
  geom_hline(yintercept = -log10(0.05), linetype = 'dashed', color = 'black')+  # Remove aes()
  geom_vline(xintercept = c(-1, 1), linetype = 'dashed', color = 'black')+  # Remove aes()
  theme_classic()+
  scale_color_manual(values = c('up_in_M'= 'blue', 'up_in_F'='salmon', 'NS'='grey'))
p1
ggsave(file.path(outpath, 'Plots', 'DiffPeaks_UBC_GNP_progenitors_MvF.pdf'), p1, width = 7, height = 5)

marker_df %>% subset(FDR < 0.05 & Log2FC > 1) %>% arrange(desc(Log2FC))
```

```{r}


p <- plotBrowserTrack(
    ArchRProj = proj1, 
    groupBy = "Sex_celltypeBroad", 
    geneSymbol = 'Maz',
    features =  getMarkers(markerPeaks, cutOff = "FDR <= 0.05 & Log2FC >= 1", returnGR = TRUE)["M_UBC_GNP_progenitors"],
    upstream = 20000,
    downstream = 20000
)

grid::grid.newpage()
grid::grid.draw(p[[1]])

```
### --- Motif enrichment for pairwise compairson
```{r}

if("Motif" %ni% names(proj1@peakAnnotation)){
    proj1 <- addMotifAnnotations(ArchRProj = proj1, motifSet = "cisbp", name = "Motif")
}

motifsUp <- peakAnnoEnrichment(
    seMarker = markerPeaks ,
    ArchRProj = proj1,
    peakAnnotation = "Motif",
    cutOff = "FDR <= 0.05 & Log2FC <= -1",
    background = "bgdPeaks"
  )
motifsUp@assays@data$Enrichment

df <- data.frame(TF = rownames(motifsUp), mlog10Padj = assay(motifsUp)[,1])
df <- df[order(df$mlog10Padj, decreasing = TRUE),]
df$rank <- seq_len(nrow(df))

ggUp <- ggplot(df, aes(rank, mlog10Padj, color = mlog10Padj)) + 
  geom_point(size = 1) +
  ggrepel::geom_label_repel(
        data = df[rev(seq_len(10)), ], aes(x = rank, y = mlog10Padj, label = TF), 
        size = 2,
        nudge_x = 2,
        color = "black",
        max.overlaps = 50
  ) + theme_ArchR() + 
  ylab("-log10(P-adj) Motif Enrichment") + 
  xlab("Rank Sorted TFs Enriched") +
  scale_color_gradientn(colors = paletteContinuous(set = "comet"))

ggUp
## Warning: ggrepel: 23 unlabeled data points (too many overlaps). Consider
## increasing max.overlaps
```

# add peak2Gene link
```{r}
# getAvailableMatrices(proj1)
# proj1 <- addPeak2GeneLinks(
#     ArchRProj = proj1,
#     reducedDims = "IterativeLSI",
#     useMatrix ="GeneExpressionMatrix"
# )
# proj1
# 
# p2g_pos <- getPeak2GeneLinks(
#     ArchRProj = proj1,
#     corCutOff = 0.01,
#     resolution = 1,
#     returnLoops = FALSE
# )
# 
# p2g_neg <- getPeak2GeneLinks(
#     ArchRProj = proj1,
#     corCutOff =-0.01,
#     resolution = 1,
#     returnLoops = FALSE
# )
# 
# p2g_pos_df <- as.data.frame(p2g_pos)
# p2g_pos_df$geneName <- mcols(metadata(p2g_pos)$geneSet)$name[p2g_pos$idxRNA]
# p2g_neg_df <- as.data.frame(p2g_neg) 
# p2g_neg_df$geneName <- mcols(metadata(p2g_neg)$geneSet)$name[p2g_neg$idxRNA]
# p2g_df <- rbind(p2g_pos_df, p2g_neg_df)
# 
# M_up_peaks <- marker_df %>% subset(FDR < 0.05 & Log2FC > 1)
# 
# M_up_peaks_geneAnot <- left_join(M_up_peaks, p2g_df, by = c('idx'='idxATAC')) %>% subset(!is.na(idxRNA))
# as.data.frame(table(M_up_peaks_geneAnot$geneName)) %>% arrange(desc(Freq))
```


```{r}

# markerGenes  <- c(
#     "IGFBP2" 
#   )
# 
# 
# 
# p <- plotBrowserTrack(
#     ArchRProj = proj1, 
#     groupBy = "rna_Sex", 
#     geneSymbol = markerGenes, 
#     upstream = 50000,
#     downstream = 50000
# )

```




# chromVar

```{r}


workDir <- file.path(Sys.getenv("COMMONR"),"sex_bias")
inpath <- file.path(workDir, "scripts/UBCenrichedE16Multiome/snATAC_mutiome") # dir with original arrow files
outpath <- file.path(workDir, "processed/mouse/multiome/snATAC/UBCenrichedE16Multiome_RNAATACshared")


set.seed(1)

cat("# Add ArchR params\n")
addArchRThreads(threads = 1) 

```

```{r}

proj1 <- loadArchRProject(path = outpath)
proj1

colnames(proj1@cellColData)

proj1@cellColData$Sex_celltypeBroad <- paste0(proj1@cellColData$rna_Sex, '_', proj1@cellColData$rna_human_celltype_pred_broad)

proj1@cellColData$Sex_celltype <- paste0(proj1@cellColData$rna_Sex, '_', proj1@cellColData$rna_human_celltype_pred)
```

### -- add chromVar matrix
```{r}

# proj1 <- addBgdPeaks(proj1)
# 
# proj1 <- addDeviationsMatrix(
#   ArchRProj = proj1, 
#   peakAnnotation = "Motif",
#   force = TRUE
# )
# 
# getAvailableMatrices(proj1) 
# saveArchRProject(ArchRProj = proj1, outputDirectory =outpath, load = FALSE)

```
# Correlate TF expression with motif accessibility
```{r}

# run once

corGSM_MM <- correlateMatrices(
    ArchRProj = proj1,
    useMatrix1 = "GeneExpressionMatrix",
    useMatrix2 = "MotifMatrix",
    reducedDims = "IterativeLSI"
)

```

### --- pairwise chromvar activity

```{r}

groupBy <- "Sex_celltypeBroad"


markerMotifs <- getMarkerFeatures(
    ArchRProj = proj1,
    useMatrix = "MotifMatrix",
    groupBy = groupBy,
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon",
    useGroups = "M_UBC_GNP_progenitors",
  bgdGroups = "F_UBC_GNP_progenitors",
  useSeqnames = 'z'
)

markerMotifs

marker_df <- data.frame(
    markerMotifs@elementMetadata
)

marker_df$FDR <-(markerMotifs@assays@data$FDR[,1])
marker_df$MeanDiff<- markerMotifs@assays@data$MeanDiff[,1]
marker_df$Mean<- markerMotifs@assays@data$Mean[,1]
marker_df$MeanBGD<- markerMotifs@assays@data$MeanBGD[,1]
marker_df <- marker_df %>% subset(MeanBGD > 0) # make sure the large difference is not due to reference as 0

```



```{r}


corGSM_MM$MeanDiff <- marker_df[match(corGSM_MM$MotifMatrix_name, marker_df$name), "MeanDiff"]
corGSM_MM$FDR <- marker_df[match(corGSM_MM$MotifMatrix_name, marker_df$name), "FDR"]

# 4. Filter for active TF regulators
corGSM_MM <- corGSM_MM[order(abs(corGSM_MM$cor), decreasing = TRUE), ]
corGSM_MM <- corGSM_MM[which(!duplicated(gsub("\\-.*", "", corGSM_MM[,"MotifMatrix_name"]))), ]

corGSM_MM$TFRegulator <- "NO"
corGSM_MM$TFRegulator[which(
    ceiling((corGSM_MM$cor) * 10) >= 5 & 
    corGSM_MM$padj < 0.05 & 
    abs(as.numeric(corGSM_MM$MeanDiff)) > 0.5 &  
    as.numeric(corGSM_MM$FDR) < 0.05
)] <- "YES"

corGSM_MM[corGSM_MM$TFRegulator=="YES", ]
corrlated_TFs <- sort(corGSM_MM[corGSM_MM$TFRegulator=="YES", 1])

corGSM_MM[corGSM_MM$GeneExpressionMatrix_name == "Etv1",]
```

```{r}

# markerMotifs <- readRDS(file.path(outpath, 'RDS', 'Sex_celltypeBroad_UBC_GNP_progenitors_markerMotifs.RDS'))
# metadata(markerMotifs)

fdr_threshold <- 0.05
diff_t <- 0.5

# Label differential genes
marker_df$label <- ifelse(marker_df$FDR < fdr_threshold & abs(marker_df$MeanDiff) > diff_t,
                          ifelse(marker_df$MeanDiff > 0, 'up_in_M', 'up_in_F'), 'NS')

# Extract gene name
marker_df$name_short <- sub('_.*', '', marker_df$name)

# Filter genes to label
is_significant <- marker_df$label != 'NS' & !grepl('ENSMUS|Gm', marker_df$name_short)
marker_df$sig_genes <- ifelse(is_significant & marker_df$name_short %in% corrlated_TFs, 
                              marker_df$name_short, NA)
marker_df$sig_genes2 <- ifelse(is_significant & marker_df$name_short %in% add_markers & 
                               !marker_df$name_short %in% corrlated_TFs, 
                               marker_df$name_short, NA)

# Plot
p1 <- ggplot(marker_df, aes(x = MeanDiff, y = -log10(FDR))) +
  geom_point(aes(color = label)) +
    geom_hline(yintercept = -log10(fdr_threshold), linetype = 'dashed', color = 'red') +
  geom_vline(xintercept = c(-diff_t, diff_t), linetype = 'dashed', color = 'red') +
  # ggrepel::geom_text_repel(aes(label = sig_genes2), max.overlaps = 20, color = 'darkgray', size = 3.5) +
  ggrepel::geom_label_repel(aes(label = sig_genes), max.overlaps =20, color = 'black', size = 3.5) +

  theme_classic() +
  scale_color_manual(values = c('up_in_M' = 'blue', 'up_in_F' = 'salmon', 'NS' = 'grey'))
p1
ggsave(file.path(outpath, 'Plots', 'DiffPeaks_UBC_GNP_progenitors_MvF.pdf'), p1, width = 6, height = 4)
unique(marker_df$sig_genes)


```
# Motif footprinting
```{r}
# library(BSgenome.Mmusculus.UCSC.mm39)
# proj1 <- addGroupCoverages(ArchRProj = proj1, groupBy = "Sex_celltypeBroad")
# proj1 <- addGroupCoverages(ArchRProj = proj1, groupBy = "Sex_celltype")
# 
# saveArchRProject(ArchRProj = proj1, outputDirectory =outpath, load = FALSE)
```


```{r}
motifPositions <- getPositions(proj1)

motifs <- c('Lmx1a', 'Otx2', 'Barhl1', 'Gbx1')
markerMotifsNames <- unlist(lapply(motifs, function(x) grep(x, names(motifPositions), value = TRUE)))
markerMotifsNames

library(BSgenome)
seFoot <- getFootprints(
  ArchRProj = proj1, 
  positions = motifPositions[markerMotifsNames], 
 groupBy = "Sex_celltypeBroad",
 useGroups = c("F_UBC_GNP_progenitors","M_UBC_GNP_progenitors", "F_UBC","M_UBC" ),
)



plotFootprints(
  seFoot = seFoot,
  ArchRProj = proj1, 
  normMethod = "Subtract",
  plotName = "Footprints-Subtract-Bias",
  addDOC = FALSE,
  smoothWindow = 10,
  pal = c("F_UBC_GNP_progenitors" = "salmon","M_UBC_GNP_progenitors"="turquoise", "F_UBC" ="maroon","M_UBC"='darkblue' )
)

# seGroupMotif <- getGroupSE(ArchRProj = proj1, useMatrix = "MotifMatrix",groupBy = "Sex_celltypeBroad")

```

```{r}

pwm <- getPeakAnnotation(proj1, "Motif")$motifs[["Otx2_437"]]

PWMatrixToProbMatrix <- function(x){
  if (class(x) != "PWMatrix") stop("x must be a TFBSTools::PWMatrix object")
  m <- (exp(as(x, "matrix"))) * TFBSTools::bg(x)/sum(TFBSTools::bg(x))
  m <- t(t(m)/colSums(m))
  m
}

ppm <- PWMatrixToProbMatrix(pwm)
ppm

library(ggseqlogo)
p1 <- ggseqlogo(ppm, method = "bits")
p1
ggsave(workDir <- file.path(Sys.getenv("COMMONR"),"sex_bias","figures/TF_motif.pdf"), width = 5, height = 2)

```
# --- closest gene to TF's motifs
```{r}

motifPositions <- getPositions(proj1)

motifs <- c('Lmx1a', 'Otx2', 'Barhl1', 'Gbx1')
markerMotifsNames <- unlist(lapply(motifs, function(x) grep(x, names(motifPositions), value = TRUE)))
markerMotifsNames

# "Lmx1a_456"  "Otx2_437"   "Barhl1_457" "Gbx1_571"  
motif_gr <- motifPositions[[ "Lmx1a_456"]]  # Use actual motif ID

# Find nearest genes using ArchR's gene annotation
gene_anno <- getGeneAnnotation(proj1)
nearest_genes <- distanceToNearest(motif_gr, gene_anno$genes, ignore.strand=FALSE)

# Extract gene names
genes <- unique(gene_anno$genes[subjectHits(nearest_genes)]$symbol)
# genes[!grepl('ENSMUSG|Gm', genes)]


```

#### --- get motif positions and matrices
```{r}

# run once
motifPositions <- getPositions(proj1)

# Load matrices once
activity_mat <- getMatrixFromProject(proj1, useMatrix = "GeneScoreMatrix")
expr_mat <- getMatrixFromProject(proj1, useMatrix = "GeneExpressionMatrix")
gene_anno <- getGeneAnnotation(proj1)

rna_obj_shared <- readRDS(Sys.getenv("COMMONR"),"sex_bias","processed/mouse/multiome/writes/UBCenrichedE16Multiome_ATACshared_seurat.RDS"))
rna_sp <- rna_obj_shared@assays$RNA$data
colnames(rna_sp) <- rna_obj_shared@meta.data$atac_cellid
expr_mat_log1p <- SummarizedExperiment(
  assays  = SimpleList(data = rna_sp),
  rowData = DataFrame(name = rownames(rna_sp)),
  colData = DataFrame(cellNames = colnames(rna_sp))
)
expr_mat_log1p 
expr_mat
```

```{r}
motif_name <- "Barhl1_457"
get_filtered_targets <- function(motif_name, proj, activity_mat, expr_mat, gene_anno, 
                                 groupby, group1, group2) {
  
  # Get motif positions and nearest genes
  motif_gr <- motifPositions[[motif_name]]
  nearest <- distanceToNearest(motif_gr, gene_anno$genes, ignore.strand = FALSE)
  genes <- unique(gene_anno$genes[subjectHits(nearest)]$symbol)
  genes <- genes[!grepl('ENSMUSG|Gm|Rik', genes)]
  length(genes)
  
  # Get genes present in both matrices
  genes_in_activity <- genes[genes %in% rowData(activity_mat)$name]
  genes_in_expr <- genes[genes %in% rowData(expr_mat)$name]
  genes_both <- intersect(genes_in_activity, genes_in_expr)
  
  # Now indices have same length
  activity_idx <- match(genes_both, rowData(activity_mat)$name)
  expr_idx <- match(genes_both, rowData(expr_mat)$name)

  # Subset cells
  group_col <- getCellColData(proj, select = groupby)[,1]
  cells_g1 <- proj$cellNames[group_col == group1]
  cells_g2 <- proj$cellNames[group_col == group2]

  
  # Rest of function same...
  activity_g1 <- rowMeans(assay(activity_mat)[activity_idx, cells_g1])
  activity_g2 <- rowMeans(assay(activity_mat)[activity_idx, cells_g2])
  
  expr_g1 <- rowMeans(assay(expr_mat)[expr_idx, cells_g1])
  expr_g2 <- rowMeans(assay(expr_mat)[expr_idx, cells_g2])
  

  # Convert to dense matrix first
  g1_mat <- as.matrix(assay(activity_mat)[activity_idx, cells_g1])
  g2_mat <- as.matrix(assay(activity_mat)[activity_idx, cells_g2])
  
  activity_test <- matrixTests::row_wilcoxon_twosample(g1_mat, g2_mat)
  activity_pval <- activity_test$pvalue
  
  # Convert to dense matrix and use vectorized test
  g1_expr <- as.matrix(assay(expr_mat)[expr_idx, cells_g1])
  g2_expr <- as.matrix(assay(expr_mat)[expr_idx, cells_g2])
  
  expr_test <- matrixTests::row_wilcoxon_twosample(g1_expr, g2_expr)
  expr_pval <- expr_test$pvalue
  
  result <- data.frame(
    TF=motif_name,
    gene = rowData(activity_mat)$name[activity_idx],
    activity_g1 = activity_g1,
    activity_g2 = activity_g2,
    activity_diff = activity_g2 - activity_g1,
    activity_pval = activity_pval,
    activity_padj = p.adjust(activity_pval, method = "BH"),
    expr_g1 = expr_g1,
    expr_g2 = expr_g2,
    expr_diff = expr_g2 - expr_g1,
    expr_log2fc = log2((expr_g2 + 1) / (expr_g1 + 1)),
    expr_pval = expr_pval,
    expr_padj = p.adjust(expr_pval, method = "BH")
  )
  
  return(result)
}


```


# --- find differentially expressed/active genes from the motif targeted genes
```{r}

motifs <- c('Lmx1a', 'Otx2', 'Barhl1')
markerMotifsNames <- unlist(lapply(motifs, function(x) grep(x, names(motifPositions), value = TRUE)))
markerMotifsNames

# M vs. F differentially expressed/active genes where lmx1a, otx2, barhl1 motif is
results_RLsvz <- do.call(rbind,lapply(markerMotifsNames, function(m) {
  get_filtered_targets(m, proj1, activity_mat, expr_mat_log1p, gene_anno,
                       groupby = "Sex_celltypeBroad",
                       group1 = "F_UBC_GNP_progenitors", 
                       group2 = "M_UBC_GNP_progenitors")
}))

results_RLsvz#%>% subset(expr_padj < 0.05 & activity_padj  < 0.05 & expr_log2fc < 0 & activity_diff <0) %>% subset(TF == "Lmx1a_456")

# write.csv(results_RLsvz, file.path(workDir, 'tables', 'RLsvz_M_vs_F_DiffOpenExp_Lmx1a_Otx2_Barhl1_motif_targeted_genes_comp.csv'))

# UBC cells
results_ubc <- do.call(rbind,lapply(markerMotifsNames, function(m) {
  get_filtered_targets(m, proj1, activity_mat, expr_mat_log1p, gene_anno,
                       groupby = "Sex_celltypeBroad",
                       group1 = "F_UBC", 
                       group2 = "M_UBC")
}))

results_ubc#%>% subset(expr_padj < 0.05 & activity_padj  < 0.05 & expr_log2fc < 0 & activity_diff <0) %>% subset(TF == "Lmx1a_456")

# write.csv(results_ubc, file.path(workDir, 'tables', 'UBC_M_vs_F_DiffOpenExp_Lmx1a_Otx2_Barhl1_motif_targeted_genes_comp.csv'))

results_RLsvz$cell_type <- 'UBC_GNP_progenitors'
results_ubc$cell_type <- 'UBC'
results <- rbind(results_RLsvz,results_ubc)

# write.csv(results, file.path(workDir, 'tables', 'RLsvz_UBC_M_vs_F_DiffOpenExp_Lmx1a_Otx2_Barhl1_motif_targeted_genes_comp.csv'))
```

```{r}
markerMotifsNames
M_up_diff_Rlsvz <- results_RLsvz %>% subset(activity_padj < 0.05 & activity_diff > 0 & expr_padj < 0.05 & expr_log2fc > 0)
M_up_diff_ubc <- results_ubc %>% subset(activity_padj < 0.05 & activity_diff > 0 & expr_padj < 0.05 & expr_log2fc > 0)
as.data.frame(table(M_up_diff_Rlsvz$gene)) %>% arrange(desc(Freq))
as.data.frame(table(M_up_diff_ubc$gene)) %>% arrange(desc(Freq))
```

```{r}

library(dplyr)
library(ComplexHeatmap)


plot_gene_cell_expr_activity <- function(df,
                                        cur_TF,
                                        proj,
                                        activity_mat,
                                        expr_mat,
                                        groupby,
                                        group1,
                                        group2,
                                        n_cells_each = NULL,   # NULL => all cells
                                        seed = 1,
                                        padj_cutoff = 0.05,
                                        max_genes = 50,
                                        scale_mode = c("zscore", "none"),
                                        clip = 2,              # 2 => [-2,2], 1 => [-1,1], NULL => no clip
                                        use_raster = TRUE,
                                        title_prefix = "") {
  set.seed(seed)
  scale_mode <- match.arg(scale_mode)

  # ---- pick genes (union across TFs), top by expr_log2fc ----
  genes_use <- df %>%
  filter(
    TF == cur_TF,
    activity_padj < padj_cutoff,
    activity_diff > 0,
    expr_padj < padj_cutoff,
    expr_log2fc > 0,
    # pmin(activity_g1 , activity_g2) > 0,
    # pmin(expr_g1 , expr_g2) > 0
  ) %>%
  mutate(score = activity_diff * expr_log2fc) %>%
  arrange(desc(score)) %>%
  slice_head(n = max_genes) %>%
  pull(gene)


  # ---- cells ----
  group_col <- getCellColData(proj, select = groupby)[, 1]
  cells_g1 <- proj$cellNames[group_col == group1]
  cells_g2 <- proj$cellNames[group_col == group2]

  cells_use <- if (is.null(n_cells_each)) {
    c(cells_g1, cells_g2)
  } else {
    c(sample(cells_g1, min(n_cells_each, length(cells_g1))),
      sample(cells_g2, min(n_cells_each, length(cells_g2))))
  }

  # ---- subset matrices ----
  g_act <- rowData(activity_mat)$name
  g_exp <- rowData(expr_mat)$name
  genes_both <- intersect(genes_use, intersect(g_act, g_exp))
  stopifnot(length(genes_both) > 1)

  idx_act <- match(genes_both, g_act)
  idx_exp <- match(genes_both, g_exp)

  common_cells <- intersect(cells_use, intersect(colnames(assay(activity_mat)), colnames(assay(expr_mat))))
  stopifnot(length(common_cells) > 1)

  M_act  <- as.matrix(assay(activity_mat)[idx_act, common_cells, drop = FALSE])
  M_expr <- as.matrix(assay(expr_mat)[idx_exp, common_cells, drop = FALSE])
  rownames(M_act)  <- genes_both
  rownames(M_expr) <- genes_both

  # order columns: group1 then group2
  col_group <- ifelse(colnames(M_expr) %in% cells_g1, group1, group2)
  ord_cols <- order(col_group)
  M_act  <- M_act[, ord_cols, drop = FALSE]
  M_expr <- M_expr[, ord_cols, drop = FALSE]
  col_group <- col_group[ord_cols]

  # ---- scaling (row-wise) + optional clipping ----
  row_z <- function(m) { m2 <- t(scale(t(m))); m2[!is.finite(m2)] <- 0; m2 }
  clip_mat <- function(m, lim) {
    if (is.null(lim)) return(m)
    dn <- dimnames(m)
    d  <- dim(m)
    m2 <- pmax(-lim, pmin(lim, m))
    dim(m2) <- d
    dimnames(m2) <- dn
    m2
}

  if (scale_mode == "zscore") {
    M_expr <- row_z(M_expr)
    M_act  <- row_z(M_act)
  }
  M_expr <- clip_mat(M_expr, clip)
  M_act  <- clip_mat(M_act,  clip)


  # ---- colors (fixed to clip range if provided) ----
  lim_expr <- if (!is.null(clip)) clip else as.numeric(quantile(abs(M_expr), 0.95, na.rm = TRUE))
  lim_act  <- if (!is.null(clip)) clip else as.numeric(quantile(abs(M_act),  0.95, na.rm = TRUE))
  col_expr <- colorRamp2(c(-lim_expr,0, lim_expr), c("blue", 'white', "red"))
  col_act  <- colorRamp2(c(-lim_act,0, lim_act),  c("purple", 'white', "#ffd22b"))

  ha <- HeatmapAnnotation(
  Group = col_group,
  col = list(
    Group = setNames(
      c("#E41A1C",  "#3a53a4"),
      c(group1, group2)
    )
  )
)

    # 1) build and draw expression heatmap ONCE to get row order
    ht_expr0 <- Heatmap(
      M_expr,
      name = "Expr",
      col = col_expr,
      top_annotation = ha,
      show_column_names = FALSE,
      cluster_rows = TRUE,
      cluster_columns = FALSE,
      use_raster = TRUE
    )
    
    ht_expr0_drawn <- draw(ht_expr0)
    ord_rows <- row_order(ht_expr0_drawn)
    
    # 2) rebuild BOTH heatmaps with fixed row order (no clustering)
    ht_expr <- Heatmap(
      M_expr[ord_rows, , drop = FALSE],
      name = "Expr",
      col = col_expr,
      top_annotation = ha,
      show_column_names = FALSE,
      cluster_rows = T,
      cluster_columns = FALSE,
      use_raster = TRUE
    )
    
    ht_act <- Heatmap(
      M_act[ord_rows, , drop = FALSE],
      name = "Activity",
      col = col_act,
      top_annotation = ha,
      show_column_names = FALSE,
      cluster_rows = T,
      cluster_columns = FALSE,
      use_raster = TRUE
    )
    
    # 3) now draw the NEW combined heatmaps (NOT ht_expr0_drawn)
    suppressMessages(draw(ht_expr + ht_act))


  invisible(list(expr = M_expr, activity = M_act, genes = genes_both, cells = colnames(M_expr), row_order = ord_rows))
}


a

```


```{r}

a <- length(proj1[proj1@cellColData$Sex_celltype == "F_UBC_GNP_progenitors"]$cellNames)
b <- length(proj1[proj1@cellColData$Sex_celltype == "M_UBC_GNP_progenitors"]$cellNames)
c <- min(a, b)
pdf(file.path(workDir, 'figures',"TF_Barhl1_expr_activity_heatmap.pdf"), width = 8, height = 5)
plot_gene_cell_expr_activity(
  df = results_RLsvz,
  cur_TF = 'Barhl1_457',
  proj = proj1,
  activity_mat = activity_mat,
  expr_mat = expr_mat_log1p,
  groupby = "Sex_celltypeBroad",
  group1 = "F_UBC_GNP_progenitors"  ,
  group2 = "M_UBC_GNP_progenitors"  ,
  n_cells_each =min(a, b),
   scale_mode ='zscore',
  max_genes = 30,
  title_prefix = "RLsvz TF targets: ",
  clip = 1
)
dev.off()

pdf(file.path(workDir, 'figures',"TF_Otx2_expr_activity_heatmap.pdf"), width = 8, height = 5)
plot_gene_cell_expr_activity(
  df = results_RLsvz,
  cur_TF = 'Otx2_437',
  proj = proj1,
  activity_mat = activity_mat,
  expr_mat = expr_mat_log1p,
  groupby = "Sex_celltypeBroad",
  group1 = "F_UBC_GNP_progenitors"  ,
  group2 = "M_UBC_GNP_progenitors"  ,
  n_cells_each =min(a, b),
   scale_mode ='zscore',
  max_genes = 30,
  title_prefix = "RLsvz TF targets: ",
  clip = 1
)
dev.off()

pdf(file.path(workDir, 'figures',"TF_Lmx1a_expr_activity_heatmap.pdf"), width = 8, height = 5)
plot_gene_cell_expr_activity(
  df = results_RLsvz,
  cur_TF = 'Lmx1a_456',
  proj = proj1,
  activity_mat = activity_mat,
  expr_mat = expr_mat_log1p,
  groupby = "Sex_celltypeBroad",
  group1 = "F_UBC_GNP_progenitors"  ,
  group2 = "M_UBC_GNP_progenitors"  ,
  n_cells_each =min(a, b),
   scale_mode ='zscore',
  max_genes = 30,
  title_prefix = "RLsvz TF targets: ",
  clip = 1
)
dev.off()
```


```{r}
library(dplyr)
library(ggVennDiagram)

padj_cutoff = 0.05
TF_ids <- c("Lmx1a_456", 'Otx2_437', 'Barhl1_457')  # replace with your 3 IDs

TGs_upM <- list(
LMX1A = results_RLsvz %>%
    filter(
      TF == TF_ids[1],
      activity_padj < padj_cutoff,
      activity_diff > 0,
      # expr_padj < padj_cutoff,
      expr_log2fc > 0
    ) %>%
  arrange(desc(activity_diff)) %>%
   # slice_head(n = 200) %>%
    pull(gene) %>%
    unique(),

OTX2 = results_RLsvz %>%
    filter(
      TF == TF_ids[2],
      activity_padj < padj_cutoff,
      activity_diff > 0,
      # expr_padj < padj_cutoff,
      expr_log2fc > 0
    ) %>%
    arrange(desc(activity_diff)) %>%
  # slice_head(n = 200) %>%
    pull(gene) %>%
    unique(),

BARHL1  = results_RLsvz %>%
    filter(
      TF == TF_ids[3],
      activity_padj < padj_cutoff,
      activity_diff > 0,
      # expr_padj < padj_cutoff,
      expr_log2fc > 0
    ) %>%
    arrange(desc(activity_diff)) %>%
   # slice_head(n = 200) %>%
    pull(gene) %>%
    unique()
)

# check sizes
sapply(TGs_upM , length)

library(VennDiagram)
library(grid)

# your list of gene sets (must be length 3)
stopifnot(length(TGs_upM) == 3)
set_sizes <- sapply(TGs_upM, length)
pdf(file.path(workDir, 'figures',"Lmx1a_TGs_upM_venn.pdf"), width = 6, height = 6)

grid.newpage()
grid.draw(
  venn.diagram(
    x = TGs_upM,
    filename = NULL,
    scaled = TRUE,
    fill = c("#89C75F", "#F37B7D", "#FEE500"),
    alpha = 0.5,
    cat.cex = 1.2,
    category.names = paste0(
      names(TGs_upM), "\n(n=", set_sizes, ")"
    )
  )
)

dev.off()


library(eulerr)

pdf(file.path(workDir, 'figures',"TGs_upM_venn.pdf"), width = 6, height = 6)

fit <- euler(TGs_upM)
plot(
  fit,
  fills = c("#89C75F", "#F37B7D", "#FEE500"),
  alpha = 0.5,
  quantities = TRUE
)

dev.off()

```
### --- ORA
```{r}
# install.packages("enrichR")
library(enrichR)
library(dplyr)
library(readr)


cur_TF <- "Barhl1_457" # "Lmx1a_456"  "Otx2_437"   "Barhl1_457" "Gbx1_571"  
cur_df <- results_RLsvz %>% subset(activity_padj < 0.05 & activity_diff > 0 &  expr_log2fc > 0 & TF == cur_TF )


genes <- unique(cur_df$gene)
length(genes)


dbs <- listEnrichrDbs()
dbs %>% arrange(desc(libraryName))
dbs_use <- c(
  "GO_Biological_Process_2025",
  # "KEGG_2019_Mouse",
  "MSigDB_Hallmark_2020",
  "WikiPathways_2019_Mouse"
)

# ---------------------------
# 3) Run Enrichr ORA
# ---------------------------
enr <- enrichr(genes, dbs_use)

# ---------- main: stack all enrichr db outputs ----------
stack_enrichr_results <- function(enr_list) {
  bind_rows(lapply(names(enr_list), function(db) {
    x <- enr_list[[db]]
    if (is.null(x) || !is.data.frame(x) || nrow(x) == 0) return(NULL)
    x %>% mutate(DB = db, .before = 1)
  }))
}

all_res <- stack_enrichr_results(enr) # function below
write.csv(all_res, file.path(workDir, 'figures', paste0(cur_TF, "_targetGenes_upM_GSEA_enrichR.csv")))
file.path(workDir, 'figures', paste0(cur_TF, "_targetGenes_upM_GSEA_enrichR.csv"))
```
### --- collapse enrichR output
```{r}
library(dplyr)
library(stringr)
library(tibble)

collapse_by_overlap_global <- function(df,
                                      genes_col   = "Genes",
                                      term_col    = "Term",
                                      padj_col    = "Adjusted.P.value",
                                      odds_col    = "Odds.Ratio",
                                      prefer_col  = "Adjusted.P.value",
                                      jaccard_max = 0.4,
                                      min_genes   = 2,
                                      keep_terms  = NULL,
                                      filter_terms = NULL) {
  # deps: dplyr, stringr
  parse_gene_set <- function(x) {
    if (is.na(x) || !nzchar(x)) return(character(0))
    strsplit(x, ";", fixed = TRUE)[[1]] |> trimws() |> (\(v) v[v != ""])()
  }
  jaccard <- function(a, b) {
    if (!length(a) || !length(b)) return(0)
    length(intersect(a, b)) / length(union(a, b))
  }

  keep_terms2   <- if (is.null(keep_terms)) character(0) else keep_terms
  filter_terms2 <- if (is.null(filter_terms)) character(0) else filter_terms
  larger_is_better <- identical(prefer_col, odds_col)

  stopifnot(all(c(genes_col, term_col, padj_col, odds_col, prefer_col) %in% colnames(df)))

  df2 <- df |>
    dplyr::filter(!( .data[[term_col]] %in% filter_terms2 )) |>
    dplyr::mutate(
      .genes = lapply(as.character(.data[[genes_col]]), parse_gene_set),
      .ng = vapply(.genes, length, integer(1)),
      .force_keep = (.data[[term_col]] %in% keep_terms2)
    ) |>
    dplyr::filter(.ng >= min_genes) |>
    dplyr::arrange(dplyr::desc(.force_keep), .data[[padj_col]], dplyr::desc(.ng))

  df2$.force_keep[is.na(df2$.force_keep)] <- FALSE

  better <- function(i_new, i_old) {
    a <- df2[[prefer_col]][i_new]
    b <- df2[[prefer_col]][i_old]
    if (is.na(a)) a <- if (larger_is_better) -Inf else Inf
    if (is.na(b)) b <- if (larger_is_better) -Inf else Inf
    if (a != b) return(if (larger_is_better) a > b else a < b)

    pn <- df2[[padj_col]][i_new]
    po <- df2[[padj_col]][i_old]
    if (!is.na(pn) && !is.na(po) && pn != po) return(pn < po)

    df2$.ng[i_new] > df2$.ng[i_old]
  }

  kept_idx <- integer(0)
  kept_sets <- list()

  for (i in seq_len(nrow(df2))) {
    gi <- df2$.genes[[i]]

    if (isTRUE(df2$.force_keep[i]) || length(kept_idx) == 0) {
      kept_idx <- c(kept_idx, i)
      kept_sets[[length(kept_sets) + 1]] <- gi
      next
    }

    overlaps <- vapply(kept_sets, function(gk) jaccard(gi, gk), numeric(1))
    hit <- which(overlaps > jaccard_max)

    if (!length(hit)) {
      kept_idx <- c(kept_idx, i)
      kept_sets[[length(kept_sets) + 1]] <- gi
    } else {
      j <- hit[which.max(overlaps[hit])]
      if (better(i, kept_idx[j])) {
        kept_idx[j] <- i
        kept_sets[[j]] <- gi
      }
    }
  }

  df2[kept_idx, , drop = FALSE] |>
    dplyr::select(-.genes, -.ng, -.force_keep)
}




```

```{r}


term_to_exclude <- all_res$Term[grepl("Cancer|UV|Adhesion|Apo|Ossi|Dea|Estro|Nucleophile|Pluri",all_res$Term)]


all_res_sig <- all_res %>% filter(Adjusted.P.value <= 0.05)

all_res %>% subset(grepl("Mitotic", Term))

collapsed <- collapse_by_overlap_global(
  all_res_sig,
  prefer_col = "Adjusted.P.value",
  jaccard_max = 0.5,  # tune this
  min_genes = 3,
  keep_terms  = c("Mitotic Spindle Organization"),
  filter_terms = term_to_exclude
  
  
)

collapsed <- collapsed %>%
  mutate(
    Term_shorten = Term %>%
      str_remove("\\s*\\(GO:\\d+\\)") %>%   # remove (GO:xxxxxxx)
      str_remove("\\s*\\(?WP\\d+\\)?") %>%  # remove WPxxxx or (WPxxxx)
      str_squish()
  )  %>%
    arrange(Adjusted.P.value) %>%
    slice_head(n = 10) %>%
    mutate(
      Term = factor(Term_shorten, levels = rev(Term_shorten)),
      neglog10FDR = -log10(Adjusted.P.value)
    )
collapsed <- collapsed %>% distinct(tolower(Term), .keep_all = TRUE) %>% 
  subset(!Term %in% c("mRNA Cis Splicing, via Spliceosome",
                      "Negative Regulation of Transcription by RNA Polymerase II", "Negative Regulation of DNA-templated Transcription", "Positive Regulation of DNA-templated Transcription"))
tf_name <- unlist(strsplit(cur_TF, split= "_", fixed = T))[1]
color_dict <- c('Lmx1a' = "#89C75F", 'Otx2' = "#F37B7D", "Barhl1"="#FEE500")
p1 <- ggplot(collapsed, aes(x = neglog10FDR, y = Term)) +
    geom_col(color = 'lightblue', fill = color_dict[tf_name]) +
    labs(x = "-log10(FDR)", y = NULL, title = ) +
    theme_classic()+
  theme(axis.text = element_text(size = 12, color = 'black'), axis.title  = element_text(size = 12, color = 'black'))
p1 

ggsave(file.path(workDir, 'figures', paste0(cur_TF, "_targetGenes_GSEA_barplots.pdf")), p1, width=10, height=5)
```

### --- check motif footprint in motifs close to MvsF diff open genes
```{r}
cur_motif <- "Lmx1a_456" #"Lmx1a_456"  "Otx2_437"   "Barhl1_457"

# 1. Get significantly different genes from your results
sig_genes <- results_RLsvz %>% subset(TF == cur_motif) %>%
  subset(activity_padj < 0.05 & activity_diff > 0) %>%
  pull(gene)

# 2. Get gene coordinates
gene_anno <- getGeneAnnotation(proj1)
sig_gene_gr <- gene_anno$genes[gene_anno$genes$symbol %in% sig_genes]

# 3. Filter motif positions - only keep motifs near significant genes
motif_gr_full <- motifPositions[[cur_motif]]

# Find motifs overlapping or near significant genes (e.g., within 5kb)
overlaps <- findOverlaps(
  motif_gr_full, 
  resize(sig_gene_gr, width(sig_gene_gr) + 3000, fix = "center"),
  ignore.strand = TRUE
)

motif_gr_filtered <- motif_gr_full[queryHits(overlaps)]
motif_list <- GRangesList("motif" = motif_gr_filtered)

library(BSgenome)
seFoot_filtered <- getFootprints(
  ArchRProj = proj1,
  positions = motif_list,
  groupBy = "Sex_celltypeBroad",
 useGroups = c("F_UBC_GNP_progenitors","M_UBC_GNP_progenitors", "F_UBC","M_UBC" ),
)

# 5. Plot
plotFootprints(
  seFoot = seFoot_filtered,
  ArchRProj = proj1,
  normMethod = "Subtract",
  plotName = "Footprints-Lmx1a-SigGenes",
  addDOC = FALSE,
  smoothWindow = 10,
  pal = c("F_UBC_GNP_progenitors" = "salmon","M_UBC_GNP_progenitors"="turquoise", "F_UBC" ="maroon","M_UBC"='darkblue')
)



```


# Save UMAP embedding
```{r}
wnnumapcoord <- read.csv(file.path(outpath, 'RDS', 'UBCEnrichedE16Multiome_SignacObject_GlutaCellsOnly_wnn.umap.coord.csv'), row.names = 1)

# Convert to matrix
wnnumapcoord_mat <- as.matrix(wnnumapcoord)

# Get all cells in proj1
proj1_cells <- rownames(proj1@cellColData)

# Create full embedding matrix with NAs
full_embedding <- matrix(NA, nrow = length(proj1_cells), ncol = 2)
rownames(full_embedding) <- proj1_cells

# Fill in cells that have embeddings
common_cells <- intersect(rownames(wnnumapcoord_mat), proj1_cells)
full_embedding[common_cells, ] <- wnnumapcoord_mat[common_cells, ]

cat("Total cells in proj1:", length(proj1_cells), "\n")
cat("Cells with embeddings:", length(common_cells), "\n")
cat("Missing embeddings:", sum(is.na(full_embedding[,1])), "\n")

# Convert to dataframe with proper column names
embedding_df <- as.data.frame(full_embedding)
colnames(embedding_df) <- paste0("UMAP_WnnUMAP_GlutaOnly#", c("UMAP_Dimension_1", "UMAP_Dimension_2"))

# Add to ArchR
proj1@embeddings$UMAP_WnnUMAP_GlutaOnly <- SimpleList(
    df = embedding_df,
    params = proj1@embeddings$UMAP_ATAC$params
)

# Plot
p <- plotEmbedding(
    ArchRProj = proj1, 
    colorBy = "MotifMatrix", 
    name = "z:Myog_45", 
    embedding = "UMAP_WnnUMAP_GlutaOnly",
    plotAs = "points"
) +facet_wrap(~proj1$rna_Sex) 
  
p

# saveArchRProject(ArchRProj = proj1, outputDirectory =outpath, load = FALSE)

```

### -- visalize in glutacell only umap
```{r}

motif_counts <- readRDS("chromvar_z.rds")
motif_dev <- saveRDS( "chromvar_deviations.rds")
motif_ranges <- saveRDS("chromvar_motifs.rds")


write.csv(as.data.frame(motif_counts), file.path(outpath, 'exportMM/chromVar', 'UBCEnrichedE16Multiome_chromVarMatrix.csv'))

# glutaOnly_obj <- readRDS(file.path(outpath, 'RDS', 'UBCEnrichedE16Multiome_SignacObject_GlutaCellsOnly.RDS'))
# wnnumapcoord <- as.data.frame(glutaOnly_obj@reductions$wnn.umap@cell.embeddings)
# write.csv(wnnumapcoord, file.path(outpath, 'RDS', 'UBCEnrichedE16Multiome_SignacObject_GlutaCellsOnly_wnn.umap.coord.csv'))

wnnumapcoord <- read.csv(file.path(outpath, 'RDS', 'UBCEnrichedE16Multiome_SignacObject_GlutaCellsOnly_wnn.umap.coord.csv'), row.names = 1)

# visualize in python scanpy object instead

```

# export peak matrix
```{r}

# write.csv(proj1@cellColData, file.path(outpath,'tables','cellMetadata.csv'), row.names = T)


# TileMatrix rownames
matrixData <- getMatrixFromProject(proj1, useMatrix = "PeakMatrix" , binarize = F)
head(matrixData[1:5,1:5])

peakRanges_df <- matrixData@assays@data$PeakMatrix

class(peakRanges_df)


# # Save the sparse matrix to a file
# writeMM(peakRanges_df, file = file.path(outpath, 'exportMM', 'PeakMatrix',"matrix.mtx")) # binarized
writeMM(peakRanges_df, file = file.path(outpath, 'exportMM', 'PeakMatrixNB',"matrix.mtx"))

# # Export cellnames
# writeLines(colnames(peakRanges_df), file.path(outpath, 'exportMM', 'PeakMatrix', "cell_names.txt"))
writeLines(colnames(peakRanges_df), file.path(outpath, 'exportMM', 'PeakMatrixNB', "cell_names.txt"))

# Export region names
row_ranges <- rowRanges(matrixData)
region_names <- paste0(seqnames(row_ranges), ":", start(row_ranges), "-", end(row_ranges))
# writeLines(region_names, file.path(outpath, 'exportMM', 'PeakMatrix', "region_names.txt"))
writeLines(region_names, file.path(outpath, 'exportMM', 'PeakMatrixNB', "region_names.txt"))


# write.csv(as.data.frame(matrixData@colData), file.path(outpath, 'exportMM', 'PeakMatrix','cell_metadata.csv'), row.names = T, col.names = T)
write.csv(as.data.frame(matrixData@colData), file.path(outpath, 'exportMM', 'PeakMatrixNB','cell_metadata.csv'), row.names = T, col.names = T)
```

#### --- extract info under seurat4 env
```{r}
workDir <- "/research/groups/northcgrp/home/common/Rachel/sex_bias"
inpath <- file.path(workDir, "scripts/UBCenrichedE16Multiome/snATAC_mutiome") # dir with original arrow files
outpath <- file.path(workDir, "processed/mouse/multiome/snATAC/UBCenrichedE16Multiome_RNAATACshared")

```

```{r}


# library(ArchR)
# library(Seurat)
# 
# 
# rna_obj_shared <- readRDS("/research/groups/northcgrp/home/common/Rachel/sex_bias/processed/mouse/multiome/writes/UBCenrichedE16Multiome_ATACshared_seurat.RDS")
# 
# # 1. Export ATAC data from ArchR
# peak_matrix <- getMatrixFromProject(proj1, useMatrix = "PeakMatrix")
# peak_counts <- assay(peak_matrix)
# peak_ranges <- rowRanges(peak_matrix)
# 
# # Save ATAC data
# saveRDS(peak_counts, "peak_counts.rds")
# saveRDS(peak_ranges, "peak_ranges.rds")

# # 1.2. Export ChromVar data from ArchR
# motif_matrix <- getMatrixFromProject(proj1, useMatrix = "MotifMatrix")
# motif_counts <- assay(motif_matrix, "z")
# motif_dev <- assay(motif_matrix, "deviations")
# motif_ranges <- rownames(motif_matrix)
# 
# # Save chromvar data
# saveRDS(motif_counts, "chromvar_z.rds")
# saveRDS(motif_dev, "chromvar_deviations.rds")
# saveRDS(motif_ranges, "chromvar_motifs.rds")

# # 1.3. Export TileMatrix data from ArchR
# tile_matrix <- getMatrixFromProject(proj1, useMatrix = "TileMatrix", binarize = T)
# tile_count <- assay(motif_matrix)
# tile_ranges <- rowRanges(motif_matrix)
# 
# # Save chromvar data
# saveRDS(tile_counts, "TileMatrix.rds")
# saveRDS(tile_ranges, "TileMatrix_ranges.rds")

# 
# # 2. Export LSI from ArchR
# lsi_matrix <- getReducedDims(proj1, reducedDims = "IterativeLSI")
# saveRDS(lsi_matrix, "lsi_embeddings.rds")
# 
# # 3. Export ATAC UMAP
# umap_atac <- getEmbedding(proj1, embedding = "UMAP_ATAC")
# saveRDS(umap_atac, "umap_atac.rds")
# 
# # 4. Export metadata
# metadata <- as.data.frame(getCellColData(proj1))
# saveRDS(metadata, "metadata.rds")
# 
# # 5. Export RNA counts from Seurat v4 object
# rna_counts <- rna_obj_shared@assays$RNA$counts
# saveRDS(rna_counts, "rna_counts.rds")
# 
# # 6. Export RNA embeddings if available
# if ("pca" %in% Seurat::Reductions(rna_obj_shared)) {
#     pca_embeddings <- Seurat::Embeddings(rna_obj_shared, reduction = "pca")
#     saveRDS(pca_embeddings, "pca_embeddings.rds")
# }
# 
# if ("umap" %in% Seurat::Reductions(rna_obj_shared)) {
#     umap_rna <- Seurat::Embeddings(rna_obj_shared, reduction = "umap")
#     saveRDS(umap_rna, "umap_rna.rds")
# }
# 
# # If you have scVI
# if ("scVI" %in% Seurat::Reductions(rna_obj_shared)) {
#     scvi_embeddings <- Seurat::Embeddings(rna_obj_shared, reduction = "scVI")
#     saveRDS(scvi_embeddings, "scvi_embeddings.rds")
# }

```
#### --- save signac object under seurat5 env
```{r}
workDir <- "/research/groups/northcgrp/home/common/Rachel/sex_bias"
inpath <- file.path(workDir, "scripts/UBCenrichedE16Multiome/snATAC_mutiome") # dir with original arrow files
outpath <- file.path(workDir, "processed/mouse/multiome/snATAC/UBCenrichedE16Multiome_RNAATACshared")

```

```{r}

library(Seurat)
library(Signac)
library(GenomicRanges)

# 1. Load all exported data
peak_counts <- readRDS("peak_counts.rds")
peak_ranges <- readRDS("peak_ranges.rds")

chromvar_motifs_mean <- readRDS("chromvar_z.rds")
chromvar_motifs_dev <- readRDS("chromvar_deviations.rds")
chromvar_motifs_ranges <- readRDS("chromvar_motifs.rds")

# tile_counts <- readRDS("TileMatrix.rds")
# tile_ranges <- readRDS("TileMatrix_ranges.rds")

lsi_matrix <- readRDS("lsi_embeddings.rds")
umap_atac <- readRDS("umap_atac.rds")
metadata <- readRDS("metadata.rds")
rna_counts <- readRDS("rna_counts.rds")
colnames(rna_counts) <- rownames(metadata)

rna_obj_shared <- readRDS("/research/groups/northcgrp/home/common/Rachel/sex_bias/processed/mouse/multiome/writes/UBCenrichedE16Multiome_ATACshared_seurat.RDS")
rna_sp <- rna_obj_shared@assays$RNA$data
colnames(rna_sp) <- rna_obj_shared@meta.data$atac_cellid
expr_mat_log1p <- SummarizedExperiment::SummarizedExperiment(
  assays  = SimpleList(data = rna_sp),
  rowData = DataFrame(name = rownames(rna_sp)),
  colData = DataFrame(cellNames = colnames(rna_sp))
)

# Optional embeddings
pca_embeddings <- readRDS("pca_embeddings.rds")
scvi_embeddings <- readRDS("scvi_embeddings.rds")
umap_rna <- readRDS("umap_rna.rds")

# 2. Create Seurat object with RNA
seurat_obj <- CreateSeuratObject(
    counts = rna_counts,
    assay = "RNA",
    meta.data = metadata
)

seurat_obj@assays$RNA$data <- rna_sp

# 3. Add ATAC assay
seurat_obj[["ATAC"]] <- CreateChromatinAssay(
    counts = peak_counts,
    ranges = peak_ranges,
    genome = "mm39"
)


# 3. Add chromvar assay
seurat_obj[["chromvar"]] <- CreateAssayObject(counts = chromvar_motifs_mean)

# 4. Add pre-computed embeddings
seurat_obj[["LSI"]] <- CreateDimReducObject(
    embeddings = lsi_matrix,
    key = "LSI_",
    assay = "ATAC"
)

seurat_obj[["umap_atac"]] <- CreateDimReducObject(
    embeddings = as.matrix(umap_atac),
    key = "UMAPATAC_",
    assay = "ATAC"
)

rownames(pca_embeddings) <- rownames(metadata)
if (exists("pca_embeddings")) {
    seurat_obj[["pca"]] <- CreateDimReducObject(
        embeddings = pca_embeddings,
        key = "PC_",
        assay = "RNA"
    )
}

rownames(scvi_embeddings) <- rownames(metadata)
if (exists("scvi_embeddings")) {
    seurat_obj[["scvi"]] <- CreateDimReducObject(
        embeddings = scvi_embeddings,
        key = "scVI_",
        assay = "RNA"
    )
}

# 5. Run multimodal analysis

# Choose which RNA reduction to use
rna_reduction <- "scvi"
seurat_obj <- FindMultiModalNeighbors(
    seurat_obj,
    reduction.list = list(rna_reduction, "LSI"),
    dims.list = list(1:50, 2:30)
)

seurat_obj <- RunUMAP(
    seurat_obj,
    nn.name = "weighted.nn",
    reduction.name = "wnn.umap",
    reduction.key = "wnnUMAP_", min_dist=0.1, spread=2
)

# 6. Visualize
DimPlot(seurat_obj, reduction = "wnn.umap", group.by = "rna_Sex")
FeaturePlot(seurat_obj, reduction = "wnn.umap",features  = "Lmx1a")

saveRDS(seurat_obj, file.path(outpath, 'RDS', 'UBCEnrichedE16Multiome_SignacObject.RDS'))

```
### --- RNA+ATAC UMAP for cell of interest only
```{r}

seurat_obj <- readRDS( file.path(outpath, 'RDS', 'UBCEnrichedE16Multiome_SignacObject.RDS'))

# Subset cells
cells_to_keep <- rownames(seurat_obj@meta.data)[
    grepl("^RL|^UBC|^GN", seurat_obj@meta.data$rna_human_celltype_pred_broad)
]
unique(seurat_obj@meta.data$rna_human_celltype_pred_broad[grepl("^RL|^UBC|^GN", seurat_obj@meta.data$rna_human_celltype_pred_broad)])



# # Rerun multimodal neighbors
# seurat_obj_subset <- FindMultiModalNeighbors(
#     seurat_obj_subset,
#     k.nn = 50,
#     reduction.list = list('scvi', "LSI"),
#     dims.list = list(1:50, 2:30)
# )
# 
# # Rerun UMAP
# seurat_obj_subset <- RunUMAP(
#     seurat_obj_subset,
#     nn.name = "weighted.nn",
#     reduction.name = "wnn.umap",
#     reduction.key = "wnnUMAP_", min_dist=0.1, spread=2
# )


seurat_obj_subset <- readRDS( file.path(outpath, 'RDS', 'UBCEnrichedE16Multiome_SignacObject_GlutaCellsOnly.RDS'))
seurat_obj_subset
umap_coords <- seurat_obj_subset@reductions$wnn.umap

seurat_obj_subset <- subset(seurat_obj, cells = cells_to_keep)

seurat_obj_subset@reductions$wnn.umap <- umap_coords

seurat_obj_subset

# Visualize

FeaturePlot(seurat_obj_subset, reduction = "wnn.umap", features = c('Lmx1a', 'Eomes', 'Mki67'))

seurat_obj_subset@meta.data$celltype <- ifelse(seurat_obj_subset@meta.data$rna_human_celltype_pred_broad == 'GNP', 'UBC_GNP_progenitors', as.character(seurat_obj_subset@meta.data$rna_human_celltype_pred_broad))
DimPlot(seurat_obj_subset, reduction = "wnn.umap", group.by = c("celltype"))

DefaultAssay(seurat_obj_subset) <- 'chromvar'
motif_names <- rownames(seurat_obj_subset@assays$chromvar$counts)
cur_motif <- motif_names[grepl('Lmx1a', motif_names)]
FeaturePlot(seurat_obj_subset, reduction = "wnn.umap", features =cur_motif, split.by = 'rna_Sex')

DefaultAssay(seurat_obj_subset) <- 'chromvar'

saveRDS(seurat_obj_subset, file.path(outpath, 'RDS', 'UBCEnrichedE16Multiome_SignacObject_GlutaCellsOnly.RDS'))
```

```{r}

workDir <- file.path(Sys.getenv("COMMONR"),"sex_bias")
inpath <- file.path(workDir, "scripts/UBCenrichedE16Multiome/snATAC_mutiome") # dir with original arrow files
outpath <- file.path(workDir, "processed/mouse/multiome/snATAC/UBCenrichedE16Multiome_RNAATACshared")


seurat_obj_subset <- readRDS( file.path(outpath, 'RDS', 'UBCEnrichedE16Multiome_SignacObject_GlutaCellsOnly.RDS'))
seurat_obj_subset
```

```{r}
library(Seurat)
library(Signac)
library(GenomicRanges)

# 1. Load all exported data
peak_counts <- readRDS("peak_counts.rds")
peak_ranges <- readRDS("peak_ranges.rds")
lsi_matrix <- readRDS("lsi_embeddings.rds")
umap_atac <- readRDS("umap_atac.rds")
metadata <- readRDS("metadata.rds")
rna_counts <- readRDS("rna_counts.rds")

# embeddings
pca_embeddings <- readRDS("pca_embeddings.rds")
scvi_embeddings <- readRDS("scvi_embeddings.rds")
umap_rna <- readRDS("umap_rna.rds")

# 2. Create Seurat object with RNA
seurat_obj <- CreateSeuratObject(
    counts = rna_counts,
    assay = "RNA",
    meta.data = metadata
)

# 3. Add ATAC assay
seurat_obj[["ATAC"]] <- CreateChromatinAssay(
    counts = peak_counts,
    ranges = peak_ranges,
    genome = "mm39"
)

# 4. Add pre-computed embeddings
seurat_obj[["LSI"]] <- CreateDimReducObject(
    embeddings = lsi_matrix,
    key = "LSI_",
    assay = "ATAC"
)

seurat_obj[["umap_atac"]] <- CreateDimReducObject(
    embeddings = umap_atac,
    key = "UMAPATAC_",
    assay = "ATAC"
)

if (exists("pca_embeddings")) {
    seurat_obj[["pca"]] <- CreateDimReducObject(
        embeddings = pca_embeddings,
        key = "PC_",
        assay = "RNA"
    )
}

if (exists("scvi_embeddings")) {
    seurat_obj[["scvi"]] <- CreateDimReducObject(
        embeddings = scvi_embeddings,
        key = "scVI_",
        assay = "RNA"
    )
}

# 5. Run multimodal analysis
# If you have scVI, use it; otherwise compute PCA
if (!"pca" %in% Reductions(seurat_obj)) {
    DefaultAssay(seurat_obj) <- "RNA"
    seurat_obj <- NormalizeData(seurat_obj)
    seurat_obj <- FindVariableFeatures(seurat_obj)
    seurat_obj <- ScaleData(seurat_obj)
    seurat_obj <- RunPCA(seurat_obj)
}

# Choose which RNA reduction to use
rna_reduction <- if ("scvi" %in% Reductions(seurat_obj)) "scvi" else "pca"

seurat_obj <- FindMultiModalNeighbors(
    seurat_obj,
    reduction.list = list(rna_reduction, "LSI"),
    dims.list = list(1:30, 2:30)
)

seurat_obj <- RunUMAP(
    seurat_obj,
    nn.name = "weighted.nn",
    reduction.name = "wnn.umap",
    reduction.key = "wnnUMAP_"
)

# 6. Visualize
DimPlot(seurat_obj, reduction = "wnn.umap", group.by = "Sex")

# 7. Save final object
saveRDS(seurat_obj, "signac_multimodal_object.rds")



```


